---
title: "Bayesian Weighted Quantile Sum Regression"
author: "Nicolo Foppa Pedretti, Elena Colicino"
date: "`r format(Sys.Date(),'%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    keep_md: true
vignette: >
  %\VignetteIndexEntry{How to use BWQS package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The Bayesian Weighted Quantile Sum (BWQS) regression is a statistical model for accommodating mixtures. The algorithm provides a weighted index of the mixture components and its overall association with an outcome, which can be continuous or binary. The overall association of the mixture index (BWQS index) is estimated adjusting for a set of covariates. The prior distributions of all parameters are considered uninformative. Parameters mapped to the overall mixture, and the covariates follow uaninformative Normal distributions, while we assumed a Dirichlet prior density for the weights. The BWQS package has several dependencies, which are uploaded automatically with the BWQS package.

The algorithms are applied to a simulated dataset and all the steps are listed in order to make reproducible analyses. In this first part of the code the environment is set up with all the useful libraries.
 

```{r setup}
library(BWQS)
library(MASS)
library(knitr)
```

## Bayesian Quantile Weighted Sums

The function `bwqs` fits the Bayesian Quantile Weighted Sums model. The object `family` allows the model to be used for both continuous or binary outcomes. The model is the following $$y_i = \beta_0 + \beta_1 \sum_{i=0}^5w_jq_{ij} + \sum_{k=0}^2 \delta_kc_{ik}$$ where $\beta_0$ is the intercept, $\beta_1$ is the overall association of the mixture with the outcome, $\sum_{i=0}^5w_jq_{ij}$ is the mixture composition made by the product between weights and quartiles (`q = 4` in `bwqs` function) of the metals concentrations.   


Synthetic dataset made of 1000 observations with a mixture composed by 5 metals (Al, As, Cd, Pb, V). The coefficients are $\beta_0 = -2$, $\beta_1 = 0.8$, the set of weights $w = [0.50,0.20,0.20,0.05,0.05]$ and the overall $\sigma = 1$. Two additional covariates are included in the model with two coefficients $\delta_1 = 0.5$ and $\delta_2 = 1.1$. Here the dataset:   

```{r example_bwqs, echo=FALSE}
# fix the seed
set.seed(1990)

# Sample sizes
N = 1000

# Mean & SD of variables
mu = c(0,0,0,1,3)
sd = c(1,1,1,3,1)
sd2 = sd %*% t(sd)

# Correlation Matrix
rho = 0.65
corMat = cbind(c(1,rho,rho^2,rho^2,rho^2),
               c(rho,1,rho^2, rho^2, rho^2),
               c(rho^2,rho^2,1,rho^2,rho^2),
               c(rho^2,rho^2,rho^2,1, rho),
               c(rho^2,rho^2,rho^2,rho,1))

# Covariance Matrix
Sigma = sd2*corMat

# Simulating five correlated exposure variables (Metals)
X = as.data.frame(mvrnorm(N, mu = mu, Sigma = Sigma, empirical = TRUE))
colnames(X) = c('Al','As','Cd','Pb','V')

# Quantile trasformation
Xq = as.matrix(quantile_split(X, mix_name = colnames(X), q=4))

# Intercept coefficient
beta0 = -2

# Overall effect
beta1 = -0.8

# Weights
W = c(0.5,0.20,0.2,0.05,0.05)

# sigma of the model
sigma = 1

# Simulate covariates
D = cbind(rnorm(N,1.3,1), rnorm(N,0.5,0.5))
colnames(D) = c("cov1","cov2")

# coefficient of covariates
delta = c(0.5, 1.1)

# Outcome simulation (continuos)
y = rnorm(N, beta0 + beta1*(Xq %*% W) + D %*% delta, sd = sigma)

# Aggregate data in a data.frame
Data = as.data.frame(cbind(y,X,D))

Data
```

Now that we have the dataset we can fit the **BWQS** model with the following code: 

```{r}
# we run the model ans save results in "fit_bwqs" variable
fit_bwqs = bwqs(y ~ cov1 + cov2, mix_name = c('Al','As','Cd','Pb','V'),
                data = Data, q = 4, family = "gaussian")
```

The first function object is the underlying model, i.e. formula `y ~ cov1 + cov2`; here we have the outcome `y` and two covariates. Additional covariates can be included in the formula separated by `+`. If no covariates are included in the model, the formula has to be specified as `y ~ NULL`. The `mix_name` object contains a **vector of strings** with the name of the mixture components. Data is the name of the dataset, containing the outcome, the mixture components, and the covariates, note that the dataset should ve in `data.frame` format. The `q` object indicates how the mixture components are categorized (as quantiles), by default is set on 4 (quartiles), but also allows deciles (`q = 10`), centiles (`q = 100`) or continuous values (`q = NULL`). Other parameters can be modified; please check the BWQS package documentation for details.   
 
Results are saved in the `fit_bwqs` object, which can be used to plot the results. In the results table: the first two columns are the mean and the standard deviation of the mean, the two following columns contain the values of the credible interval at 95%. The credible interval level can be chosen using the `c_int` parameter. The last two columns report values to evaluate the fitness of parameters and the convergence of the algorithm. 
 
```{r tables}
fit_bwqs$summary_fit
```

The following plots summarize the results: estimates and 95% credible intervals. The first object of the function contains the results of the model. `parms` is a vector with the name of the variables to be plotted: `beta0` and `beta1` for the main parameters of the model, `W` for the weights and `delta` for the covariates. The `size` identifies the magnitude of the means dots.

```{r plots}
bwqs_plot(fit_bwqs, parms = "W", size = 2)
```

```{r plots1}
bwqs_plot(fit_bwqs, parms = c("beta0","beta1",'delta',"sigma"), size = 2)
```

The `bwqs_waic` function evaluates the goodness of fit for the BWQS model. Several measures are included in the function. For further details check [here](http://www.stat.columbia.edu/~gelman/research/unpublished/loo_stan.pdf).  

```{r waic}
bwqs_waic(fit_bwqs$fit)
```



